### 2019-02-09 ic-journal/coarsest.tex
Sync htex with tex files, clear all DS and DDS blocks

%As bisimilarity ($\approx$) is not congruence, for this reason rooted bisimilarity has been
%introduced (Def.~\ref{d:rootedBisimilarity}).
In this subsection we discuss two proofs of \DDS{the fundamental  result} stating that
rooted bisimilarity is the coarsest congruence contained in
bisimilarity \cite{van2005characterisation,Gorrieri:2015jt,Mil89}:
\begin{equation}
\label{eq:coarsest}
\forall p\ \ q.\ p\ \rapprox\ \! q\ \Longleftrightarrow\ ( \forall r.\ p\ +\
r\ \approx\ q\ +\ r )\enspace.
\end{equation}

@@

\DDS{The coarsest congruence
contained in (weak) bisimilarity, namely  \emph{bisimilarity
  congruence}, is}
the \emph{composition closure} (\texttt{CC}) of (weak) bisimilarity:
\begin{alltt}
\HOLConst{WEAK_CONGR} \HOLSymConst{=} \HOLConst{CC} \HOLConst{WEAK_EQUIV}\hfill{[WEAK_CONGR]}
\HOLConst{CC} \HOLFreeVar{R} \HOLSymConst{=} (\HOLTokenLambda{}\HOLBoundVar{g} \HOLBoundVar{h}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{c}. \HOLConst{CONTEXT} \HOLBoundVar{c} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{R} (\HOLBoundVar{c} \HOLBoundVar{g}) (\HOLBoundVar{c} \HOLBoundVar{h}))\hfill{[CC_def]}
\end{alltt}
 (We need not  put $R\ g\ h$ in the antecedent of
  \texttt{CC\_def}, as this is anyhow obtained from the trivial context $(\lambda x.\,x)$.)
\DDS{For any relation $R$ 
on CCS processes, the composition closure of $R$ is always at least as
fine as }
$R$
(here $\subseteq_r$ stands for \emph{relational subset})
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R}. \HOLConst{CC} \HOLBoundVar{R} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R}\hfill{[CC_is_finer]}
\end{alltt}
Furthermore, we prove that any (pre)congruence contained in $R$ (which
itself may not be) is contained in the composition closure of $R$
(hence the closure is the coarsest one):
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R} \HOLBoundVar{R\sp{\prime}}. \HOLConst{congruence} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLConst{CC} \HOLBoundVar{R}\hfill{[CC_is_coarsest]}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R} \HOLBoundVar{R\sp{\prime}}. \HOLConst{precongruence} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLConst{CC} \HOLBoundVar{R}\hfill{[CC_is_coarsest']}
\end{alltt}

@@

\DDS{The formal proofs  of the above two results  follow Milner's proof
\cite{Mil89}. The assumption is Milner's result 
requires  $\mathrm{fn}(p) \cup
\mathrm{fn}(q) \neq \mathscr{L}$ (where $\mathrm{fn}$ stands for \emph{free
  names}), and is therefore strictly stronger than the assumption in
our result above, where we only look at immediate weak
derivatives $p$ and $q$, requiring that there is at least one action
that never appears as a label; moreover such an action need not be
the same for $p$ and for $q$
%  If only $p$ (or $q$) has free actions while the other uses all available
% labels, the assumption $\mathrm{fn}(p) \cup
% \mathrm{fn}(q) \neq \mathscr{L}$ (here $\mathrm{fn}$ stands for \emph{free
%   names}) does not hold, and the proof cannot be completed. Our
% assumption is \emph{weaker}
% in the sense that $p$ and $q$ need not have the
% \emph{same} free actions 
(and where an action $a$ and its complementary  $\outC a$ are
considered \emph{different} actions).}

There exists a different, more complex, proof of (\ref{eq:coarsest}),
given by van Glabbeek \cite{van2005characterisation}, which does not
require any additional assumption.
The core lemma says, for
any two processes $p$ and $q$, if there exists a \emph{stable} (i.e.~$\tau$-free)
 process $k$ which is not bisimilar with any derivative of $p$ and
 $q$, then \texttt{SUM_EQUIV} indeed implies rooted bisimilarity ($\rapprox$):
\begin{alltt}
\HOLTokenTurnstile{} (\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}.
        \HOLConst{STABLE} \HOLBoundVar{k} \HOLSymConst{\HOLTokenConj{}} (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p\sp{\prime}} \HOLBoundVar{u}. \HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenConj{}}
        \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{q\sp{\prime}} \HOLBoundVar{u}. \HOLFreeVar{q} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenImp{}}
   (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}) \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{q}\hfill{[PROP3_COMMON]}
\end{alltt}
\begin{alltt}
\HOLConst{STABLE} \HOLFreeVar{p} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{u} \HOLBoundVar{p\sp{\prime}}. \HOLFreeVar{p} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{u} \HOLSymConst{\HOLTokenNotEqual{}} \HOLSymConst{\ensuremath{\tau}}\hfill{[STABLE]}
\end{alltt}
To construct this process $k$,
 the proof relies on arbitrary infinite sums of 
processes and uses transfinite induction to obtain
an arbitrary large sequence of processes (firstly introduced by Jan
Willem Klop \cite{van2005characterisation})
 that are all pairwise non-bisimilar.
We have partially formalised
this proof, because the typed logic
implemented in various HOL systems (including Isabelle/HOL) is not
strong enough to define a type for all possible ordinals
\cite{norrish2013ordinals}, thus
we have replaced transfinite induction with plain induction. As a
consequence, the final
result is about a restricted class of processes (which we have taken
to be the finite-state processes).
\DDS{The proof, in the section below, uses extensively HOL's
\texttt{pred_set} theory \cite{melham1992hol}, with a mix
of CCS and pure mathematics.}

@@

The assumption in the  previous \texttt{PROP3_COMMON} requires the
existence of a special CCS process, which is not weakly bisimilar to
any sub-process emanating from the two root processes by weak
transitions. In the worst case, there may be infinite such
sub-processes (even on finitely branching processes).
We can however consider  the equivalence classes of CCS processes 
modulo weak bisimilarity.
If there are infinitely many of such classes, 
then it will be 
possible to choose one  that is distinct from all the (finitely many) states in the
graphs of the two given processes.  
\DDS{This can be done following Klop's contruction \cite{van2005characterisation}.}
 We call  ``Klop processes''  the processes in this construction:
\begin{definition}[Klop processes]
For each ordinal $\lambda$, and an arbitrary chosen non-$\tau$ action $a$,
define a CCS process $k_\lambda$ as follows:
\begin{itemize}
\item $k_0 = 0$,
\item $k_{\lambda+1} = k_\lambda + a.k_\lambda$ and
\item for $\lambda$ a limit ordinal, $k_\lambda = \sum_{\mu < \lambda}
  k_\mu$, meaning that $k_\lambda$ is constructed from all graphs
  $k_\mu$ for $\mu < \lambda$ by identifying their root.
\end{itemize}
\end{definition}
When processes are finite-state, that is,
the number of  states in which a process may evolve by performing
transitions is finite, 
we can use  the following subset of Klop processes, 
defined as a recursive function (on natural numbers) in HOL4:
\begin{definition}{(Klop processes as recursive function on natural numbers)}
\begin{alltt}
\HOLConst{KLOP} \HOLFreeVar{a} \HOLNumLit{0} \HOLSymConst{=} \HOLConst{nil}
\HOLConst{KLOP} \HOLFreeVar{a} (\HOLConst{SUC} \HOLFreeVar{n}) \HOLSymConst{=} \HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n} \HOLSymConst{\ensuremath{+}} \HOLConst{label} \HOLFreeVar{a}\HOLSymConst{..}\HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n}\hfill{[KLOP_def]}
\end{alltt}
\end{definition}

@@

% (processes 
% all states coming from
% two finite-state CCS processes $p$ and $q$ is finite. Choosing from an
% infinite set for an element distinct with any subprocess leading from
% $p$ and $q$, is always possible.  This result is purely mathematical,
% completely falling into basic set theory:
\begin{lemma}
Given an equivalence relation $R$ defined on a type, and two sets $A, B$
of elements in this type, 
\DDS{if $A$ is finite, $B$ is infinite, and all elements
in $B$ belong to distinct equivalence classes,} then there exists an element $k$ in $B$
which is not equivalent to any element in $A$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{equivalence} \HOLFreeVar{R} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{FINITE} \HOLFreeVar{A} \HOLSymConst{\HOLTokenConj{}} \HOLConst{INFINITE} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}}
   (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{x} \HOLBoundVar{y}. \HOLBoundVar{x} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{x} \HOLSymConst{\HOLTokenNotEqual{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{x} \HOLBoundVar{y}) \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}. \HOLBoundVar{k} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{n}. \HOLBoundVar{n} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{A} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{n} \HOLBoundVar{k}\hfill[INFINITE_EXISTS_LEMMA]
\end{alltt}
\end{lemma}
% \begin{proof}
%   We built an explicit mapping $f$ from $A$ to $B$\footnote{There're
%     multiple ways to prove this lemma, a simpler proof is to make a
%     reverse mapping from $B$ to the power set of $A$ (or further use
%     the Axiom of Choice (AC) to make a mapping from $B$ to $A$), then
%     the non-injectivity of this mapping will contradict the fact that
%     all elements in the infinite set are distinct. Our proof doesn't
%     need AC, and it relies on very simple truths about sets.}, for all
%   $x \in A$, $y = f(x)$ if $y \in B$ and $y$ is equivalent with
%   $x$. But it's possible that no element in $B$ is equivalent with
%   $x$, and in this case we just choose an arbitrary element as
%   $f(x)$. Such a mapping is to make sure the range of $f$ always fall
%   into $B$.

@@

Using all the above results, now we can  prove the following finite
version of ``Klop lemma'':
\begin{lemma}{(Klop lemma,  finite version)}
\label{lem:klop-lemma-finite}
\DDS{For any two finite-state CCS $p$ and $q$, there is another process
$k$, which is not weakly bisimilar 
with
 any weak derivative  of $p$
and $q$ (i.e., any process reachable from $p $ or $q$ by means of transitions):}
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}.
       \HOLConst{finite_state} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}.
           \HOLConst{STABLE} \HOLBoundVar{k} \HOLSymConst{\HOLTokenConj{}} (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{p} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenConj{}}
           \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{q\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{q} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})\hfill{[KLOP_LEMMA_FINITE]}
\end{alltt}
\end{lemma}

>>>>

%As bisimilarity ($\approx$) is not congruence, for this reason rooted bisimilarity has been
%introduced (Def.~\ref{d:rootedBisimilarity}).
In this subsection we discuss two proofs of the fundamental result stating that
rooted bisimilarity is the coarsest congruence contained in
bisimilarity \cite{van2005characterisation,Gorrieri:2015jt,Mil89}:
\begin{equation}
\label{eq:coarsest}
\forall p\ \ q.\ p\ \rapprox\ \! q\ \Longleftrightarrow\ ( \forall r.\ p\ +\
r\ \approx\ q\ +\ r )\enspace.
\end{equation}

@@

The coarsest congruence
contained in (weak) bisimilarity, namely \emph{bisimilarity
  congruence}, is
the \emph{composition closure} (\texttt{CC}) of (weak) bisimilarity:
\begin{alltt}
\HOLConst{WEAK_CONGR} \HOLSymConst{=} \HOLConst{CC} \HOLConst{WEAK_EQUIV}\hfill{[WEAK_CONGR]}
\HOLConst{CC} \HOLFreeVar{R} \HOLSymConst{=} (\HOLTokenLambda{}\HOLBoundVar{g} \HOLBoundVar{h}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{c}. \HOLConst{CONTEXT} \HOLBoundVar{c} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{R} (\HOLBoundVar{c} \HOLBoundVar{g}) (\HOLBoundVar{c} \HOLBoundVar{h}))\hfill{[CC_def]}
\end{alltt}
 (We need not  put $R\ g\ h$ in the antecedent of
  \texttt{CC\_def}, as this is anyhow obtained from the trivial context $(\lambda x.\,x)$.)
For any relation $R$ 
on CCS processes, the composition closure of $R$ is always at least as
fine as $R$
(here $\subseteq_r$ stands for \emph{relational subset})
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R}. \HOLConst{CC} \HOLBoundVar{R} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R}\hfill{[CC_is_finer]}
\end{alltt}
Furthermore, we prove that any (pre)congruence contained in $R$ (which
itself may not be) is contained in the composition closure of $R$
(hence the closure is the coarsest one):
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R} \HOLBoundVar{R\sp{\prime}}. \HOLConst{congruence} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLConst{CC} \HOLBoundVar{R}\hfill{[CC_is_coarsest]}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{R} \HOLBoundVar{R\sp{\prime}}. \HOLConst{precongruence} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLBoundVar{R} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{R\sp{\prime}} \HOLSymConst{\HOLTokenRSubset{}} \HOLConst{CC} \HOLBoundVar{R}\hfill{[CC_is_coarsest']}
\end{alltt}

@@

The formal proofs  of the above two results  follow Milner's proof
\cite{Mil89}. The assumption is Milner's result 
requires  $\mathrm{fn}(p) \cup
\mathrm{fn}(q) \neq \mathscr{L}$ (where $\mathrm{fn}$ stands for \emph{free
  names}), and is therefore strictly stronger than the assumption in
our result above, where we only look at immediate weak
derivatives $p$ and $q$, requiring that there is at least one action
that never appears as a label; moreover such an action need not be
the same for $p$ and for $q$
%  If only $p$ (or $q$) has free actions while the other uses all available
% labels, the assumption $\mathrm{fn}(p) \cup
% \mathrm{fn}(q) \neq \mathscr{L}$ (here $\mathrm{fn}$ stands for \emph{free
%   names}) does not hold, and the proof cannot be completed. Our
% assumption is \emph{weaker}
% in the sense that $p$ and $q$ need not have the
% \emph{same} free actions 
(and where an action $a$ and its complementary  $\outC a$ are
considered \emph{different} actions).

There exists a different, more complex, proof of (\ref{eq:coarsest}),
given by van Glabbeek \cite{van2005characterisation}, which does not
require any additional assumption.
The core lemma says, for
any two processes $p$ and $q$, if there exists a \emph{stable} (i.e.~$\tau$-free)
 process $k$ which is not bisimilar with any derivative of $p$ and
 $q$, then \texttt{SUM_EQUIV} indeed implies rooted bisimilarity ($\rapprox$):
\begin{alltt}
\HOLTokenTurnstile{} (\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}.
        \HOLConst{STABLE} \HOLBoundVar{k} \HOLSymConst{\HOLTokenConj{}} (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p\sp{\prime}} \HOLBoundVar{u}. \HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenConj{}}
        \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{q\sp{\prime}} \HOLBoundVar{u}. \HOLFreeVar{q} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenImp{}}
   (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}) \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{q}\hfill{[PROP3_COMMON]}
\end{alltt}
\begin{alltt}
\HOLConst{STABLE} \HOLFreeVar{p} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{u} \HOLBoundVar{p\sp{\prime}}. \HOLFreeVar{p} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{u} \HOLSymConst{\HOLTokenNotEqual{}} \HOLSymConst{\ensuremath{\tau}}\hfill{[STABLE]}
\end{alltt}
To construct this process $k$,
 the proof relies on arbitrary infinite sums of 
processes and uses transfinite induction to obtain
an arbitrary large sequence of processes (firstly introduced by Jan
Willem Klop \cite{van2005characterisation})
 that are all pairwise non-bisimilar.
We have partially formalised
this proof, because the typed logic
implemented in various HOL systems (including Isabelle/HOL) is not
strong enough to define a type for all possible ordinals
\cite{norrish2013ordinals}, thus
we have replaced transfinite induction with plain induction. As a
consequence, the final
result is about a restricted class of processes (which we have taken
to be the finite-state processes).
The proof, in the section below, uses extensively HOL's
\texttt{pred_set} theory \cite{melham1992hol}, with a mix
of CCS and pure mathematics.

@@

The assumption in the  previous \texttt{PROP3_COMMON} requires the
existence of a special CCS process, which is not weakly bisimilar to
any sub-process emanating from the two root processes by weak
transitions. In the worst case, there may be infinite such
sub-processes (even on finitely branching processes).
We can however consider  the equivalence classes of CCS processes 
modulo weak bisimilarity.
If there are infinitely many of such classes, 
then it will be 
possible to choose one  that is distinct from all the (finitely many) states in the
graphs of the two given processes.  
This can be done following Klop's contruction \cite{van2005characterisation}.
 We call  ``Klop processes''  the processes in this construction:
\begin{definition}[Klop processes]
For each ordinal $\lambda$, and an arbitrary chosen non-$\tau$ action $a$,
define a CCS process $k_\lambda$ as follows:
\begin{itemize}
\item $k_0 = 0$,
\item $k_{\lambda+1} = k_\lambda + a.k_\lambda$ and
\item for $\lambda$ a limit ordinal, $k_\lambda = \sum_{\mu < \lambda}
  k_\mu$, meaning that $k_\lambda$ is constructed from all graphs
  $k_\mu$ for $\mu < \lambda$ by identifying their root.
\end{itemize}
\end{definition}
When processes are finite-state, that is,
the number of  states in which a process may evolve by performing
transitions is finite, 
we can use  the following subset of Klop processes, 
defined as a recursive function (on natural numbers) in HOL4:
\begin{definition}{(Klop processes as recursive function on natural numbers)}
\begin{alltt}
\HOLConst{KLOP} \HOLFreeVar{a} \HOLNumLit{0} \HOLSymConst{=} \HOLConst{nil}
\HOLConst{KLOP} \HOLFreeVar{a} (\HOLConst{SUC} \HOLFreeVar{n}) \HOLSymConst{=} \HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n} \HOLSymConst{\ensuremath{+}} \HOLConst{label} \HOLFreeVar{a}\HOLSymConst{..}\HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n}\hfill{[KLOP_def]}
\end{alltt}
\end{definition}

@@

% (processes 
% all states coming from
% two finite-state CCS processes $p$ and $q$ is finite. Choosing from an
% infinite set for an element distinct with any subprocess leading from
% $p$ and $q$, is always possible.  This result is purely mathematical,
% completely falling into basic set theory:
\begin{lemma}
Given an equivalence relation $R$ defined on a type, and two sets $A, B$
of elements in this type, 
if $A$ is finite, $B$ is infinite, and all elements
in $B$ belong to distinct equivalence classes, then there exists an element $k$ in $B$
which is not equivalent to any element in $A$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{equivalence} \HOLFreeVar{R} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{FINITE} \HOLFreeVar{A} \HOLSymConst{\HOLTokenConj{}} \HOLConst{INFINITE} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}}
   (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{x} \HOLBoundVar{y}. \HOLBoundVar{x} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{x} \HOLSymConst{\HOLTokenNotEqual{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{x} \HOLBoundVar{y}) \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}. \HOLBoundVar{k} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{n}. \HOLBoundVar{n} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{A} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{n} \HOLBoundVar{k}\hfill[INFINITE_EXISTS_LEMMA]
\end{alltt}
\end{lemma}
% \begin{proof}
%   We built an explicit mapping $f$ from $A$ to $B$\footnote{There're
%     multiple ways to prove this lemma, a simpler proof is to make a
%     reverse mapping from $B$ to the power set of $A$ (or further use
%     the Axiom of Choice (AC) to make a mapping from $B$ to $A$), then
%     the non-injectivity of this mapping will contradict the fact that
%     all elements in the infinite set are distinct. Our proof doesn't
%     need AC, and it relies on very simple truths about sets.}, for all
%   $x \in A$, $y = f(x)$ if $y \in B$ and $y$ is equivalent with
%   $x$. But it's possible that no element in $B$ is equivalent with
%   $x$, and in this case we just choose an arbitrary element as
%   $f(x)$. Such a mapping is to make sure the range of $f$ always fall
%   into $B$.

@@

Using all the above results, now we can  prove the following finite
version of ``Klop lemma'':
\begin{lemma}{(Klop lemma,  finite version)}
\label{lem:klop-lemma-finite}
For any two finite-state CCS $p$ and $q$, there is another process
$k$, which is not weakly bisimilar 
with
 any weak derivative  of $p$
and $q$ (i.e., any process reachable from $p $ or $q$ by means of transitions):
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}.
       \HOLConst{finite_state} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}.
           \HOLConst{STABLE} \HOLBoundVar{k} \HOLSymConst{\HOLTokenConj{}} (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{p} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenConj{}}
           \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{q\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{q} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})\hfill{[KLOP_LEMMA_FINITE]}
\end{alltt}
\end{lemma}
